---
title: "Flow Statistical Modeling: Pre-RA Aim 3 Longitudinal Study"
author: "Nicole Tran"
date: "2023-11-29"
output:
  html_document:
    toc: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

# Cluster Level Analysis

**Goal**:  Define the cellular and molecular features that change during the progression from ACPA+ autoimmunity to clinically-apparent disease


**Model**: CLR ~ time + sex + age + BMI + (1|subject)


*Adjustments*: correct adj pvalue calculations, female samples only, days to conversion <=750, remove sex as term



## Set Up
```{r, setup}
`%notin%` <- Negate(`%in%`)

outersect <- function(x, y) {
  sort(c(setdiff(x, y),
         setdiff(y, x)))
}

library(tidyverse)
library(ggpubr)
library(readxl)
library(stats)
suppressMessages(library('ggplot2'))
suppressMessages(library('data.table'))
suppressMessages(library('reshape2'))
suppressMessages(library('readr'))
suppressMessages(library('purrr'))
suppressMessages(library('plotly'))
suppressMessages(library('DT'))
library(broom)
library(lmerTest)
library(broom.mixed)

# define the color palette to be used
npg_color <- c("#E64B35FF", "#4DBBD5FF", "#00A087FF", "#3C5488FF", "#F39B7FFF", 
               "#8491B4FF", "#91D1C2FF", "#DC0000FF", "#7E6148FF", "#B09C85FF")
nejm_color <- c("#BC3C29FF", "#0072B5FF", "#E18727FF", "#20854EFF", "#7876B1FF", "#6F99ADFF", "#FFDC91FF", "#EE4C97FF")
jama_color <- c("#374E55FF", "#DF8F44FF", "#00A1D5FF", "#B24745FF", "#79AF97FF", "#6A6599FF", "#80796BFF")
jco_color <- c("#0073C2FF", "#EFC000FF", "#868686FF", "#CD534CFF", "#7AA6DCFF", "#003C67FF", "#8F7700FF")
cluster_colors <- c("#DC050C", "#FB8072", "#1965B0", "#7BAFDE", "#882E72", "#B17BA6", "#FF7F00", "#FDB462", "#E7298A", 
                    "#E78AC3", "#33A02C", "#B2DF8A", "#55A1B1", "#8DD3C7", "#A6761D", "#E6AB02", "#7570B3", "#BEAED4", "#666666", "#999999", 
                    "#aa8282", "#d4b7b7", "#8600bf", "#ba5ce3", "#808000", "#aeae5c", "#1e90ff", "#00bfff", "#56ff0d", "#ffff00")
cluster_colors_ext <- colorRampPalette(cluster_colors)(40)


################################################################################
#### functions ####
################################################################################
# Defnie a function to do clr transformatin from frequency table
FreqClrTran <- function(freq_table, freq_col='frequency_live', celltype_col='labels', sample_col='sample_id'){
    require('compositions')
    freq <- freq_table %>% select(all_of(c(freq_col, celltype_col, sample_col))) %>% 
        pivot_wider(id_cols = .data[[sample_col]], names_from =.data[[celltype_col]],
                    values_from = .data[[freq_col]])
    freq_mx <- freq %>% select(-.data[[sample_col]]) %>% as.matrix()
    rownames(freq_mx) <- freq[[sample_col]]
    freq_clr <- compositions::clr(freq_mx) %>% as_tibble(rownames = sample_col) %>% 
        pivot_longer(cols=c(-.data[[sample_col]]), names_to = celltype_col, values_to = 'clr') 
    freq_meta_clr <- freq_table %>% full_join(freq_clr, by = c(sample_col,celltype_col))
    return(freq_meta_clr)
}


# run linear mixed model for time to conversion 
freq_lme <- function(freq_table, formula){
    #message(paste('running', labels))
    lme_res <- tidy(lmer(as.formula(formula), data=freq_table)) 
    return(lme_res)
}


################################################################################
#### params ####
################################################################################

meta_aim_path <-  "~/Library/CloudStorage/GoogleDrive-nicole.tran@alleninstitute.org/Shared drives/Imm - Data/Immunology Projects/PID-00005 Autoimmunity/PLAN-00118 Pre-RA Longitudinal Analysis/Data and Analysis/Metadata/ALTRA_RA_Aim3_ALTRA_converters_longitudinal_scrna_metadata.csv"

meta_aim_bmi_path <- "~/Library/CloudStorage/GoogleDrive-nicole.tran@alleninstitute.org/Shared drives/Imm - Data/Immunology Projects/PID-00005 Autoimmunity/PLAN-00118 Pre-RA Longitudinal Analysis/Data and Analysis/Metadata/2023-11-22_ALTRA_Metadata_labs_long-model-vars.csv"

# create the directory to save figures and output
results_path <- '~/Library/CloudStorage/GoogleDrive-nicole.tran@alleninstitute.org/Shared drives/Imm - Data/Immunology Projects/PID-00005 Autoimmunity/PLAN-00118 Pre-RA Longitudinal Analysis/Data and Analysis/flow cytometry/statistical-testing/results/current/'

data_path <- '~/Library/CloudStorage/GoogleDrive-nicole.tran@alleninstitute.org/Shared drives/Imm - Data/Immunology Projects/PID-00005 Autoimmunity/PLAN-00118 Pre-RA Longitudinal Analysis/Data and Analysis/flow cytometry/statistical-testing/data/current/'

# directory of raw counts 
freq_file_paths <- "../../raw-data/leiden-cluster-freq-tabls-current/"

if(!dir.exists(results_path)){dir.create(results_path,recursive = TRUE)}
if(!dir.exists(data_path)){dir.create(data_path,recursive = TRUE)}
options(repr.plot.width = 20, repr.plot.height = 15)

aim <- 'Aim3'


clr_dir <-paste0(data_path, "clr-freq/")
if(!dir.exists(clr_dir)){dir.create(clr_dir,recursive = TRUE)}

da_dir <-paste0(data_path, "da-results/")
if(!dir.exists(da_dir)){dir.create(da_dir,recursive = TRUE)}


```

# Read in

```{r}
aim <- 'Aim3'
method <- 'GLMM'
form <- 'clr ~ days_to_conversion  + age_conv + bmi_conv + (1|subject.subjectGuid)'


################################################################################
#### metdata ####
################################################################################
meta_aim <- read_csv(meta_aim_path)

# select sample specific to aim from meta_aim
meta_aim <- meta_aim %>% select(sample.sampleKitGuid,)

str(meta_aim)

meta_aim %>% distinct(sample.sampleKitGuid)

# bmi metadata contains samples from all aims and modalities 
meta_aim_bmi <- read_csv(meta_aim_bmi_path)

meta_aim_bmi <- meta_aim_bmi %>% select(sample.sampleKitGuid,subject.subjectGuid, days_to_conversion,subject.biologicalSex,age_conv,bmi_conv, Status_Long)

# merge bmi and conversion meta with aim3 sample meta
meta_aim <- meta_aim %>% inner_join(meta_aim_bmi)

# as of Feb 15 2024, remove samples greater than -750 days
meta_aim <- meta_aim %>% filter(days_to_conversion >= -750) 

# as of Feb 16 2024, remove all male samples, remove biological sex term form
meta_aim <- meta_aim %>% filter(subject.biologicalSex %in% "Female")

################################################################################
#### automate aim  per panel ####
################################################################################
freq_tbl_path <- list.files(freq_file_paths, pattern = "*pseudocount.csv",include.dirs = T, full.names = T)

analysis <- gsub(".*select_(.*)\\_pseudocount.csv", "\\1", freq_tbl_path)


```


# Statistical Results


```{r, results='asis'}
################################################################################
#### aim params ####
################################################################################
aim <- 'Aim3'
method <- 'GLMM'
form <- 'clr ~ days_to_conversion  + age_conv + bmi_conv + (1|subject.subjectGuid)'


da_res_df <- list()
clr_res_df <- list()

for(i in seq_along(analysis)){

  print(analysis[i])
  
  panel <- str_extract(analysis[i],pattern = 'P.\\d')
  
  freq_tbl <- read_csv(freq_tbl_path[i]) 
  
  # create sample.sampleKitGuid column
  #freq_tbl$sample.sampleKitGuid <- paste0("KT",gsub("^PB(.*)\\-\\d\\d","\\1", freq_tbl$sample_id))
  
  freq_tbl$batch <- panel
  
   # remove duplicated sample and label levels
  freq_tbl <- freq_tbl %>% distinct(sample.sampleKitGuid, cluster, .keep_all = T)
  
  freq_tbl$labels <- paste(freq_tbl$cluster,analysis[i],sep = "_")
  
  # # Rename columns starting with "downsample_"
  # freq_tbl <- freq_tbl %>%
  #   rename_with(~ "cell_type_total", starts_with("downsample_"))
  
  # merge metadata for aim1 with frequency table, select only aim 1 samples
  freq_tbl_aim_meta <- freq_tbl %>% inner_join(meta_aim)
  
  ################################################################################
  #### perform CLR transformation ####
  ################################################################################
  freq_tbl_aim_meta_clr <- FreqClrTran(freq_tbl_aim_meta,freq_col = "frequency_live_pseudocount")
  
  
  clr_res_df[[analysis[i]]] <- freq_tbl_aim_meta_clr
  
 
  ################################################################################
  #### GLM ####
  ################################################################################

  freq_stats <- freq_tbl_aim_meta_clr %>%
        group_by(labels) %>% 
      group_modify(~freq_lme(.x, formula=form)) %>% 
      ungroup()

  
  # calculate the pval adj grouped by covariate
  freq_stats <- freq_stats %>% group_by(term) %>%
    rstatix::adjust_pvalue( method = "BH") %>% ungroup()
  
  
  # calculate number of patients per cell type/feature
  n_patients_per_feature <- freq_tbl_aim_meta_clr %>% group_by(labels) %>% 
    summarise(N = n_distinct(subject.subjectGuid))
  
  da_res_df[[i]] <- freq_stats %>% mutate(Aim = aim,
                            Analysis = analysis[i],
                            Method = method,
                            Formula = form) %>%
    inner_join(n_patients_per_feature) %>%
    rename(CellType = labels,
           Covariate = term)
  
}

clr_freq <- rbindlist(clr_res_df)
da_res <- rbindlist(da_res_df)

## add sig symbol to df of only the positive covariate
da_res <- da_res%>%
  filter(Covariate %notin% c("sd__(Intercept)", "sd__Observation")) %>%
rstatix::add_significance(p.col = 'p.value.adj', output.col = 'sig',
                          cutpoints = c(0, 0.05, 0.2, 1),
                          symbols = c( "*", "<0.2", "ns"))%>%
ungroup()

################################################################################
#### export results ####
################################################################################
clr_dir <-paste0(data_path, "clr-freq/")
if(!dir.exists(clr_dir)){dir.create(clr_dir,recursive = TRUE)}

write_csv(clr_freq, paste0(clr_dir, "pre-ra_flow_freq_tbl_clr_trans_cluster_res_recluster_v3_", aim, ".csv"))

da_dir <-paste0(data_path, "da-results/")
if(!dir.exists(da_dir)){dir.create(da_dir,recursive = TRUE)}

write_csv(da_res, paste0(da_dir, "pre-ra_flow_DA_statistical_results_cluster_res_pseudocount_750_DTC_female_only_recluster_v3_", aim, ".csv"))


```

# Plot 

```{r}
aim_results_path <- paste0(results_path, aim, "/")
if(!dir.exists(aim_results_path)){dir.create(aim_results_path,recursive = TRUE)}

aim <- 'Aim3'
method <- 'GLMM'
form <- 'clr ~ days_to_conversion + subject.biologicalSex + age_conv + bmi_conv + (1|subject.subjectGuid)'

da_res <- read_csv(paste0(da_dir, "pre-ra_flow_DA_statistical_results_cluster_res_pseudocount_750_DTC_female_only_recluster_v3_Aim3.csv"))

## add sig symbol to df of only the positive covariate
da_res <- da_res%>%
  filter(Covariate %notin% c("sd__(Intercept)", "sd__Observation")) %>%
rstatix::add_significance(p.col = 'p.value.adj', output.col = 'sig',
                          cutpoints = c(0, 0.05, 0.2, 1),
                          symbols = c( "*", "<0.2", "ns"))%>%
ungroup()

# reorder covar levels
da_res <- da_res %>% mutate(Covariate = factor(Covariate, levels = c("(Intercept)",
                                                           "days_to_conversion",
                                                           "subject.biologicalSexMale",
                                                           "age_conv",
                                                           "bmi_conv")))

clr_freq <- read_csv(paste0(data_path, "pre-ra_flow_freq_tbl_clr_trans_cluster_res_recluster_v3_Aim3.csv"))

clr_freq <- clr_freq %>% filter(!is.na(sample.sampleKitGuid))

analysis <- da_res$Analysis %>% unique()

for(i in seq_along(analysis)){
  proj_name <- paste0(aim, "_", analysis[i])

  
  #curr_panel <- str_extract(analysis[i], "P.\\d")

  da_df <- da_res %>% filter(Analysis %in% analysis[i]) %>%
    filter(Covariate=='days_to_conversion') %>%
    filter(p.value.adj<0.2) %>%
    arrange(p.value.adj) %>%
    mutate(CellType = as.factor(CellType))
  
  if(nrow(da_df) == 0){
    next
  }else{ 
     p <- clr_freq %>% filter(labels %in% unique(da_df$CellType)) %>%
      #filter(panel == curr_panel) %>%
      ggplot(aes(x=days_to_conversion, y=`clr`)) +
        geom_line(aes(color=factor(subject.subjectGuid))) +
      geom_point(aes(color=factor(subject.subjectGuid)))+
        stat_smooth(color = "#FC4E07", fill = "#FC4E07",
      method = "lm", alpha=0.2) + 
       facet_wrap(vars(labels), scales = 'free_y') +
      ggtitle(analysis[i],
                subtitle = paste(unique(da_df$Aim),  "\n",
                                 "Formula:", unique(da_df$Formula), "\n",
                                 "Adjust Pvalue < 0.2", "\n",
                                 "days_to_conversion >= -750 & Female Only"))
  
  
    ggsave(plot = p,
           file.path(aim_results_path, paste0(proj_name, '_sig_glmm_frequency_clr.png')),
           width=18, height=8)
    }

 


  # ### plot pvalue distributions for all covariates
  # h <- da_res %>%
  #  filter(Analysis %in% analysis[i]) %>%
  #     #filter(Covariate=='days_to_conversion') %>%
  #   ggplot(., aes(x=p.value.adj,color = Covariate, fill=Covariate)) +
  #   geom_histogram(bins = 25)+
  #   geom_vline(xintercept = .2, linetype="dotted",
  #                 color = "red", size=1) +
  #   facet_wrap(~`CellType`) +
  #   ggtitle(analysis[i],
  #               subtitle = paste(unique(da_df$Aim),  "\n",
  #                                "Formula:", unique(da_df$Formula)))
  # 
  #   ggsave(plot = h,
  #        file.path(results_path, paste0(proj_name, '_adj.p_values_distribution.png')),
  #        width=15, height=10)
  # 
  #   # volcano plot
  #   v <- da_res %>%
  #   filter(Analysis %in% analysis[i]) %>%
  #   ggplot(aes(x=estimate, y= (-log(p.value.adj)),
  #                              color=sig, shape = Covariate)) +
  #   geom_point() +
  #   facet_wrap(vars(CellType), scales = 'free_y') + geom_vline(xintercept = 0)+
  #   ggtitle(analysis[i], 
  #             subtitle = paste(unique(da_df$Aim),  "\n",
  #                              "Formula:", unique(da_df$Formula)))
  #   
  #   ggsave(plot = v,
  #        file.path(results_path, paste0(proj_name, '_covariate_volcano_plot.png')), 
  #        width=15, height=10)
  #   
  #   
  #   da_res %>%
  #   filter(Analysis %in% analysis[i]) %>%
  #   ggplot(aes(x=estimate, y= (-log(p.value.adj)),
  #                              color=CellType, shape = sig)) +
  #   geom_point() +
  #   #facet_wrap(vars(CellType), scales = 'free_y') + 
  #   geom_vline(xintercept = 0)+
  #   ggtitle(analysis[i], 
  #             subtitle = paste(unique(da_df$Aim),  "\n",
  #                              "Formula:", unique(da_df$Formula)))
  #   
}




```


# Merge Cluster with Cell type annotations

```{r}
aim = "Aim3"
annotations <- read_csv("../data/current/FC_Leiden_summary_cluster_IDs.csv")

# split unique label column
annotations <- annotations %>% separate_wider_delim(unique_label, delim = "---", names = c("Analysis", "CellType"))

################################################################################
#### DA ####
################################################################################

da_res <- read_csv(paste0(data_path, "pre-ra_flow_DA_statistical_results_cluster_res_pseudocount_750_DTC_female_only_", aim, ".csv"))

# inner merge da results with annotations
da_res_merged <- da_res %>% left_join(annotations) 

da_res_merged$labels <- da_res_merged$CellType

write_csv(da_res_merged, file = paste0(data_path, "annotations/", "pre-ra_flow_DA_statistical_results_cluster_res_pseudocount_750_DTC_female_only_annotations_", aim, ".csv"))

################################################################################
#### CLR ####
################################################################################
clr_freq <- read_csv(paste0(data_path, "pre-ra_flow_freq_tbl_clr_trans_cluster_res_pseudocount_750_DTC_female_only_", aim, ".csv"))

clr_freq$Analysis <- paste0('Flow_DA_',clr_freq$panel)


# inner merge da results with annotations
clr_freq_merged <- clr_freq %>% left_join(annotations,
                                          by = c("labels" = "CellType",
                                                 "Analysis" = "Analysis"))

clr_freq_merged$Aim <- aim

write_csv(clr_freq_merged, file = paste0(data_path, "annotations/", "pre-ra_flow_freq_tbl_clr_trans_cluster_res_750_DTC_female_only_annotations_", aim, ".csv"))



```







